---
title: "The Tidyverse - ggplot"
---

```{r setup, include=FALSE, message=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      out.width = "100%",
                      fig.align = 'center')
```
# The Tidyverse - ggplot

## The Bigger Picture

In this document we learn how to create sophisticated plots with ggplot. Simply put, we are learning how to transform tidy data into visually clear graphs. In the overall context of the workflow, this falls into the category of transforming our data into data visualisations.

<center><img src = "images/workflowschematic.JPG" width="80%" class="ImageBorder"> </img></center>
\  
<center><img src = "images/workflow.JPG" width="80%" class="ImageBorder"> </img></center>

`{{<expand "Note: LinkedIn Learning videos" "...">}}
There are references to LinkedIn Learning videos. These are complementary but not really required as the notes below are meant to be self-contained. Some students and staff would have access for free. Do not purchase access unless you are sure you don’t have access through your organisation already.
{{</expand>}}`

## What is ggplot?

> [LindedIn Learning 1.1](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/introducing-ggplot2)

```{r library-tidyverse, echo = TRUE, message = FALSE}
library("tidyverse")
```

```{r sample_data, echo=FALSE}
sample_data <- as.tibble(matrix(c(1,2,3,4,1,3,4,2), ncol = 2))
```

- A chart-making package in R
- A subset of the Tidyverse, the highly regarded data manipulating package of R
- The package is used to create highly tweakable graphs, plots and charts of many varieties
    - Scatterplots and line graphs
    - Bar graphs and column graphs
    - Histograms
    - Maps
    - And many more
- ggplot heavily depends on the add (`+`) operator to create charts

## The Grammar of Graphics

> [LindedIn Learning 1.2](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/the-grammar-of-graphics)

The 'Grammar of Graphics' is just a method of describing graphs we create. This is more relevant to `ggplot` than other packages we may use because of the unique syntax required by `ggplot`. 

For every visualisation we wish to create, we must consider its properties.

```{r bar_v_scatter, echo=FALSE}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2)) +
  xlab("x-value") +
  ylab("y-value")

sample_data %>%
  ggplot() +
  geom_col(aes(x = V1,
                 y = V2)) +
  xlab("x-value") +
  ylab("y-value")
```


In this example, both graphs display the same data, but bar graph is very different from a scatterplot. In the bar graph, the size of a bar represents the y-value of our data, whereas in the scatterplot this is represented by the height of a point. We need a language to describe what is going on here.

We describe components of a visualisation as such:

| *Component of Visualisation* | *Definition* | *Examples* |
| :- | :- | :- |
| **Data** | The set of data we want to visualise | Height, weight, number of... |
| **Geometries** | Shapes we use to visualise data | Lines, bubbles, regions... |
| **Aesthetics** | Properties of geometries | Thickness, size, colour... |
| **Scales** | Mappings between geometries and aesthetics | Thickness of a line, size of a bubble, colour of a region... |

Now we have the language we need to fully describe visualisations. One uses a point-and-line geometry, with vertical and horizontal aesthetics reflecting x- and y-values, and one uses a bar geometry, describing y-values using the height aesthetic and x-values using the label aesthetic.

## ggplot() and the "add" (`+`) Operator

> [LindedIn Learning 1.4](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/build-your-first-visualization)

### `ggplot()`

- Whenever we wish to make a visualisation with `ggplot` we begin with `ggplot()`
- This creates an empty plot

```{r ggplot()1}
ggplot()
```

- When we plot data, we note that the `ggplot()` function has a `data` argument

We can add data in two ways:

- Directly using the argument:
```{r ggplot()2, eval=FALSE}
ggplot(data = sample_data)
```
- Using the pipeline operator of the Tidyverse:
```{r ggplot()3}
sample_data %>%
  ggplot()
```

We note that no data is showing up. This is because we haven't specified any method of graphing using the "add" (`+`) operator.

### The "add" (`+`) Operator

- `ggplot` works by creating a plot with `ggplot()` and then 'add'ing to it
- We specify *data* with `ggplot()`
- We then specify *geometries* by 'add'ing other functions to `ggplot()`
- Then we can specify *aesthetics* using these functions
- As an example, if we 'add' the `geom_point()` function to our ggplot using `+`, we will get a plot

```{r add1}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2))
```

- The "add" (`+`) operator represents adding a geometry
- If we wish to add another geometry, we can just use it again
- For example, if we want lines connecting these points, we can just add another geometry with the `geom_line()` function

```{r add2}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2))
```

## Creating Scatterplots

> [LindedIn Learning 2.1](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/scatterplots)

We will be using the [ACORN-SAT temperature data](o	http://www.bom.gov.au/climate/data/acorn-sat/#tabs=Data-and-networks) for our examples.

```{r acorn-sat}
load("tidy_ACORN-SAT_data/station_data.rdata")
sample_temperature <- station_data %>%
  filter(Station.name == "Sydney"
         | Station.name == "Darwin"
         | Station.name == "Adelaide"
         | Station.name == "Alice Springs")
head(sample_temperature, 5)
```

We already know we choose **data** to plot using `ggplot()`. We know we can choose **geometries** using various other functions. Now we need to know how to specificy **aesthetics**.

- Every geometry function takes an argument called `mapping`
- This argument takes the value of another function, called `aes()`
- `aes()` itself has arguments which control our aesthetics

For a scatterplot our geometry function is `geom_point()`, `aes()` has these arguments:

| Argument | Description |
| :-: | :-: |
| x | Data for the x-axis |
| y | Data for the y-axis |
|shape | Shape of points |
| color | Colour of points |
| size | Size of points |

There is also an argument `alpha` which controls the opacity of points, *but this is an argument of `geom_point()`, **not** `aes()`*

- For our graph, we can use this to plot temperature over time

```{r scatterplot1}
sample_temperature %>%
  ggplot() +
  geom_point(aes(x = year,
                 y = average.temp))
```

- Currently the plot looks rather disorganised - we will fix this later
- For now we will add an aesthetic and modify the colour of points according to location

```{r scatterplot2}
sample_temperature %>%
  ggplot() +
  geom_point(aes(x = year,
                 y = average.temp,
                 color = Station.name))
```

## Creating Line Plots and Smoothing
> [LindedIn Learning 2.2](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/lines-and-smoothers)

- `geom_line()` is much like `geom_point()`, but it is used to create line graphs

```{r line1}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp))
```

Did the plot not work?

- To use `geom_line()` with multiple categories, we must tell ggplot how to group them somehow
- We use the `group` argument of `aes()`
- Here we specify to group temperature data according to the station it comes from

```{r line2}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name))
```

- `geom_smooth()` is a new geometry function
- This function puts a curve of best fit in our graph

```{r smooth1}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name))
```

- We can remove the grey area representing standard error of our COBF by setting the `se` argument to false

```{r smooth2}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name),
              se = FALSE)
```

## A Note on Aesthetics

So far we have seen the `aes()` function as an argument of 'geom' functions, and we will continue to see this throughout this material. It is worth noting, however, that the `aes()` function can be used as an argument of `ggplot()` *instead* of the geom functions. As an example consider this plot with multiple geometries:

```{r note1}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2))
```

This can be rewritten using only one use of `aes()` if we use it as an argument of `ggplot()` instead of both `geom_point()` and `geom_line()`:

```{r note2}
sample_data %>%
  ggplot(aes(x = V1,
             y = V2)) +
  geom_point() +
  geom_line()
```

We get the same effect. Note that if we then used `aes()` within `geom_point()` (for example), the new aesthetics we supply would override the `ggplot()` aesthetics.

## Creating Bar and Column Graphs
> [LindedIn Learning 2.3](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/bars-and-columns)


- There are two types of bar/column graph geometry functions we deal with:
    - `geom_bar()`
    - `geom_col()`

### `geom_bar()`
    
- For the `geom_bar()` function, the user specifies an `x` variable to graph, and the function takes the `y` variable to be a count of observations of the `x`
- This is similar to a histogram, only we usually use it for categorical variables
- Let's demonstrate by classifing years in Sydney as hot, cold or moderate

```{r bar1}
hot_or_cold <- station_data %>%
  filter(Station.name == "Sydney") %>%
  mutate(warmth = ifelse(average.temp > 18, "hot", "cold"))
```

- We can then see how many of each day there are

```{r bar2}
hot_or_cold %>%
  ggplot() +
  geom_bar(aes(x = warmth))
```

### `geom_col()`

- The `geom_col()` function is more versatile, allowing the user to specify both and `x` and `y` variable
- The `x` is usually categorical, and the `y` quantative

We introduce the [Australian Environmental-Economic Accounts (2016)](https://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/4655.0Explanatory%20Notes12016?OpenDocument) for our examples.

```{r envacc}
load("tidy_EnvAcc_data/consumption.rdata")
head(consumption, 12)
```

```{r bar3}
consumption %>%
  ggplot() +
  geom_col(aes(x = year,
               y = water_consumption))
```

- Here we plot Australia's total yearly water consumption
- If we wish to create a stacked column graph, we can try specifing the `color` argument of `aes()` to colour by state

```{r bar4}
consumption %>%
  ggplot() +
  geom_col(aes(x = year,
               y = water_consumption,
               color = State))
```

- This default, however, gives only an outline
- We can use the `fill` argument to create solid colour

```{r bar5}
consumption %>%
  ggplot() +
  geom_col(aes(x = year,
               y = water_consumption,
               fill = State))
```

## Creating Histograms
> [LindedIn Learning 2.4](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/histograms)

- Histograms in `ggplot` are very flexible
- We use the `geom_histogram()` function to specify a histogram geometry
- We can then fine tune several arguments (**Note**: these are arguments of `geom_histogram()`, and *not* aes())
    - The `x` argument selects the variable to plot
    - The `bins` argument can choose the number of bins we use *OR*
    - The `binwidth` argument can specify the width of the bins we place our observations in
    - The `origin` argument specifies the number from which we begin setting bins and is only useful with the `bins` argument
    
We again use annual temperatures in Sydney as an example.

- First try setting exactly 10 bins

```{r histogram1}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                 bins = 10)
```

- We can see that our first bin begins slightly before 17 degrees
- If we wished, we could use `origin` to ensure the first bin begins counting from 16.5 degrees

```{r histogram2}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                 bins = 10,
                 origin = 16.5)
```


- We could also specify a binwidth width of 0.5 degrees, meaning one column represents all observations in a 0.5 degree range

```{r histogram3}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5)
```

## Creating Boxplots
> [LindedIn Learning 2.5](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/box-plots)

- We use the `geom_boxplot()` geometry function
- The user specifies an `x` and a `y` variable
    - `x` must be a categorical variable
- This will automatically construct a boxplot for every different category of `x` based on the distribution of our `y` data

We can, as an example, examine how the average yearly temperatures of Sydney, Darwin, Adelaide and Alice Springs.

```{r boxplot1}
sample_temperature %>%
  ggplot() +
  geom_boxplot(aes(x = Station.name,
                   y = average.temp,
                   fill = Station.name))
```

## The Precision of ggplot

As was stated before, `ggplot` is a package which allows for extremely detailed tweaking of graphs. This includes the ability to modify, create and delete:

- The background colour of charts
- The grid lines thickness and colour
- The axis labels
- The axis scale
- The colours of any geometries
- The legend title
- The legend position
- Annotations
- Lines
- Titles and subtitles

We make such modifications using the `theme()` function. This function can take many different arguments:

- `plot.background`
- `panel.background`
- `panel.grid.major`
- `panel.grid.minor`
- `legend.key`
- `axis.ticks`
- `axis.title`

Each relates to one element of a visualisation, and we consider each in turn.

## Modifying Background Elements
> [LindedIn Learning 3.1](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/modifying-the-background)

For our tweaking examples, we'll use our histogram:

```{r background1, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5)
```

We introduce the `theme()` function and discover we can modify several elements.

### The Plot Background colour

- We use `plot.background` and set this to the `element_rect()` function
- This new function has the `colour` argument which we can control, making the border a colour of our choice

```{r background2, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(plot.background = element_rect(colour = "red"))
```

- Or the `fill` argument, which fills the entire background

```{r background3, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(plot.background = element_rect(fill = "red"))
```

### The Panel Background Colour

- We use `panel.background` and set this to the `element_rect()` function
- Then again we have our options to `colour` or `fill`

```{r panel1, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.background = element_rect(colour = "red"))
```

Or...

```{r panel2, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.background = element_rect(fill = "red"))
```

### Grid Lines

- We can also modify the grid lines we see on our charts
- There are two types of grid lines, major and minor
- They are modified using `panel.grid.major` and `panel.grid.minor`
- For each, we set them to the `element_line()` function
- This new function takes arguments of `colour` and `size`

Here are some examples:

- Modifying major lines to be blue and very thick:

```{r gridlines1, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.grid.major = element_line(colour = "blue",
                                        size = 5))
```

- Note: we can modify the vertical and horizontal major grid lines with `panel.grid.major.x` and `panel.grid.major.y` respectively

```{r gridlines2, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.grid.major.x = element_line(colour = "blue"),
        panel.grid.major.y = element_line(colour = "green"))
```

- Modifying minor lines:

```{r gridlines3, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.grid.minor = element_line(colour = "red"))
```

- If we wish, we can also totally remove an element (such as the grid lines)
- If we don't want any gridlines, we set it to the `element_blank()` function

```{r gridlines4, out.width="40%"}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

### Get Creative!

Combining all of these features can produce some cool looking graphs!
[If you have html colour codes](https://htmlcolorcodes.com/), you can pick your own colours!

```{r creative1}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme(plot.background = element_rect(colour = "#000000",
                                       fill = "#A6FEFD"),
        panel.background = element_rect(colour = "#000000",
                                        fill = "#FFEEEE"),
        panel.grid.major = element_line(size = 0.5,
                                        colour = "#FF8383"),
        panel.grid.minor = element_line(colour = "#B0FCFF"))
```

## Modifying the Axes
> [LindedIn Learning 3.2](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/working-with-axes)

- We introduce two new types of functions
- `xlab()` and `ylab()` can change the labels of axes
- `xlim()` and `ylim()` can change the range of what the axes display

It's simple enough to change axis names:

```{r axis1}
station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  xlab("Average Yearly Temp. (°C)") +
  ylab("Number of Observations (1910-2012")
```

As an example of axis range, observe contrast in the x-axis for the following:

```{r axis2}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2))
```

Versus...

```{r axis3}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2)) +
  xlim(0,5) +
  ylim(0,5)
```

We can also reduce the axis range, but this may cause a loss of data.

```{r axis4}
sample_data %>%
  ggplot() +
  geom_point(aes(x = V1,
                 y = V2)) +
  geom_line(aes(x = V1,
                y = V2)) +
  xlim(0,3) +
  ylim(0,3)
```

## Changing Scale
> [LindedIn Learning 3.3](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/changing-scales)

Finally, we can take care of this awfully scaled plot!

```{r scale1, echo=FALSE}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  xlab("Yearly Temp. (°C)") +
  ylab("Year")
```

We introduce the concept of scale functions.

- There are very many, each with their own purpose, but they all generally modify the scales of our charts
- They are named by the convention "scale_name_datatype()"
- A [full list can be found here](https://ggplot2.tidyverse.org/reference/#section-scales), but examples include:
    - `scale_x_discrete()`
    - `scale_y_continuous()`
    - `scale_size_continuous()`
    - `scale_fill_manual()`
    - `scale_colour_gradient()`
    - Etc...
- The arguments they take vary according to the function, but for now we are interested in:

|||
| :-: | :-: |
|`name`| Changes axis names (just like `xlab()`)|
|`limits`|Changes the axis range with precise control|
|`breaks`| Change the way numbers are displayed on our scale with precise control|

\  

First of all, we rename axes with `name` and rescale our range.
**Note:** `limits` and `breaks` take a vector of numbers as their value. We ususally use `seq()` to assign to `breaks` (specifying the start, end, and step-value) whilst we usually use a vector to assign to `limits` (specifying only a start and end value).

```{r scale2}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  scale_x_discrete(name = "Year",
                   breaks = seq(1910, 2012, 10)) +
  scale_y_continuous(name = "Yearly Temp. (°C)",
                     limits = c(15, 30))
```

Now this looks much nicer!

## Custom Colours and Cleaner Legends
> [LindedIn Learning 3.4](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/cleaning-up-legends)

- Sometimes automatically generated legends aren't exactly what we want
- We can add the `scale_fill_manual()` or `scale_colour_manual()` functions to our plot to fine tune it
    - Which do we use? If we colour our data using the `fill` argument, use the first, and if we colour our data using the `colour` argument, use the second
- This function has two important arguments -- `values` and `guide`
- The `values` function takes a vector of colours and can be used to create custom colours in a chart
    
```{r legends1}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  scale_x_discrete(name = "Year",
                   breaks = seq(1910, 2012, 10)) +
  scale_y_continuous(name = "Yearly Temp. (°C)",
                     limits = c(15, 30)) +
  scale_colour_manual(values = c("#FF0000", "#0031FF", "#FF00EB", "#13BB00"))
```
    
### Functions within functions

The `guide` argument of the above two functions itself takes a function as its value. The function is called `guide_legend()`.

- These are the arguments of `guide_legend()`:

|||
| :-: | :-: |
|`title`| Changes legend title|
|`nrow`|Changes how many rows the legend uses to display data|
|`label.position`| Changes where the label title appears in the legend|
|`keywidth`| Numeric argument to modify the width of legend boxes|
|`legend.key`| Takes value of a vector of colours and specifies custom colours|

\  

Let's use all of the above on our chart.

- We set the title to "Location"
- We display the data in two rows
- We put the legend label (here the location name) to the left
- We specify a width of 2.5

```{r legends2}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  scale_x_discrete(name = "Year",
                   breaks = seq(1910, 2012, 10)) +
  scale_y_continuous(name = "Yearly Temp. (°C)",
                     limits = c(15, 30)) +
  scale_colour_manual(values = c("#FF0000", "#0031FF", "#FF00EB", "#13BB00"),
                    guide = guide_legend(title = "Location",
                                         nrow = 2,
                                         label.position = "top",
                                         keywidth = 2.5))
```

- We note we can also use the `legend.position` argument of the `theme()` function to change where our legend is

```{r legends3}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  scale_x_discrete(name = "Year",
                   breaks = seq(1910, 2012, 10)) +
  scale_y_continuous(name = "Yearly Temp. (°C)",
                     limits = c(15, 30)) +
  scale_colour_manual(values = c("#FF0000", "#0031FF", "#FF00EB", "#13BB00"),
                    guide = guide_legend(title = "Location",
                                         nrow = 1,
                                         label.position = "top",
                                         keywidth = 2.5)) +
  theme(legend.position = "top")
```

**Note:** `legend.position= "none"` removes a legend entirely.

## Annotations
> [LindedIn Learning 3.5](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/annotating-your-visualization)

- To annotate we use the `annotate()` function
- **Note:** the first argument of this function must be "text"
- It also takes the `label` argument, being the text of your annotation
- In addition, the `x` and `y` arguments for position on a chart

```{r annotations1}
sample_temperature %>%
  ggplot() +
  geom_line(aes(x = year,
                y = average.temp,
                color = Station.name,
                group = Station.name)) +
  geom_smooth(aes(x = year,
                  y = average.temp,
                  color = Station.name,
                  group = Station.name)) +
  scale_x_discrete(name = "Year",
                   breaks = seq(1910, 2012, 10)) +
  scale_y_continuous(name = "Yearly Temp. (°C)",
                     limits = c(15, 30)) +
  scale_colour_manual(values = c("#FF0000", "#0031FF", "#FF00EB", "#13BB00")) +
  theme(legend.position = "none") +
  annotate("text",
           label = "Adelaide",
           x = 90,
           y = 16.5) +
  annotate("text",
           label = "Sydney",
           x = 90,
           y = 18.5) +
  annotate("text",
           label = "Alice Springs",
           x = 90,
           y = 22) +
  annotate("text",
           label = "Darwin",
           x = 90,
           y = 27.5)
```

- We can also add lines in as we please using the `geom_hline()` or `geom_vline()` functions
- These only take the `yintercept` or `xintercept` functions respectively and can be used to place lines at any point on a chart

```{r annotate2}
consumption %>%
  ggplot() +
  geom_col(aes(x = year,
               y = water_consumption,
               fill = State)) +
  ylim(0,22000) +
  annotate("text",
           label = "Peak",
           x = 5,
           y = 21000) +
  geom_hline(yintercept = 19756)
```

## Titles
> [LindedIn Learning 3.6](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/adding-titles)

- For titles we use `ggtitle()`
- For subtitles we use the `subtitle` argument of `ggtitle()`

```{r titles1}
consumption %>%
  ggplot() +
  geom_col(aes(x = year,
               y = water_consumption,
               fill = State)) +
  ggtitle("Water Consumption of Australia over Time",
          subtitle = "Data provided by the Australian Environmental-Economic Accounts, 2016")
```

## Predefined Themes
> [LindedIn Learning 3.7](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/using-themes)

- In as much as we have precise customisation available in ggplot, sometimes we may not want to bother
- We can automatically modify graphs we've made using pre-constructed themes from ggplot, using set theme functions as such

```{r themes1}
library("ggthemes")

station_data %>%
  filter(Station.name == "Sydney") %>%
  ggplot() +
  geom_histogram(aes(x = average.temp),
                     binwidth = 0.5) +
  theme_solarized()
```

- Some themes included in standard `ggplot`:
    - `theme_bw()`
    - `theme_dark()`
    - `theme_void()`
    - `theme_minimal()`
- Some themes included in the `ggthemes` package (must be installed and library'd)
    - `theme_solarized()`
    - `theme_excel()`
    - `theme_wsi()`
    - `theme_economist()`
    - `theme_fivethirtyeight()`
- [Here are the ggplot themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), and [here are the ggthemes themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/)

## A Useful Link

Here's the [ggplot Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) published by R Studio!


## Mapping with ggplot and ggmap
> [LindedIn Learning Section 4](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/visualizing-data-with-maps)

- `ggplot` and its sibling package `ggmap` are used to render maps and map-based charts in R
- This includes the ability to make, among other things:
    - Regular maps
    - Geoplots
    - Choropleths
- ggmap communicates with Google Maps, and functions by using an API key (Application Programming Interface key) to request access to Google's maps

Unfortunately, as of June 2018, Google updated its policy on API keys, significantly limiting the capabilities of ggmap for individuals who have not signed up to Google APIs with billing details. As a result the content of Section 4 is not considered in this course.

## ggplot Challenge
> [LindedIn Learning 5.1](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2/challenge-assignment)

The LindedIn Learning tutorial provides a practice dataset, along with a challenge. Given the dataset, the challenge is:

- Plot a map of all schools in California
- Colour the points based on their Institutional Control
- Change the size of the points to reflect their Undergraduate Population

Good luck!