---
title: "htmlwidgets - DT"
---

```{r setup, include=FALSE, message=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      out.width = "80%")
```

# htmlwidgets - DT

## The Bigger Picture

In this document we learn how to create interactive tables with DT. Simply put, we are learning how to transform tidy data into visually clear tables. In the overall context of the workflow, this falls into the category of transforming our data into data visualisation.

<center><img src = "images/workflowschematic.JPG" width="80%" class="ImageBorder"> </img></center>
\  
<center><img src = "images/workflow.JPG" width="80%" class="ImageBorder"> </img></center>

`{{<expand "Note: LinkedIn Learning videos" "...">}}
There are references to LinkedIn Learning videos. These are complementary but not really required as the notes below are meant to be self-contained. Some students and staff would have access for free. Do not purchase access unless you are sure you don’t have access through your organisation already.
{{</expand>}}`

## What is DT?

> [LinkedIn Learning 6.1](https://www.linkedin.com/learning/r-interactive-visualizations-with-htmlwidgets/what-is-the-dt-library)

```{r library-highcharter, echo = TRUE, message = FALSE}
library("tidyverse")
library("DT")
```

- An htmlwidget used to make interactive data tables from data frames and tibbles
- It is arguably the best R package for this specific purpose
- The package is bound to the DataTable library in JavaScript

## Creating Interactive Tables
> [LinkedIn Learning 6.2](https://www.linkedin.com/learning/r-interactive-visualizations-with-htmlwidgets/create-interactive-data-tables-with-dt)

Let's say we have a dataset, such as the set below from the [ACORN-SAT dataset](http://www.bom.gov.au/climate/data/acorn-sat/#tabs=Data-and-networks). Note that we are deliberately using a large dataset (this one contains around 6000 observations)

```{r create1}
load("tidy_ACORN-SAT_data/station_data.rdata")
head(station_data, 8)
```

With DT, we can instantly make this an interactive table by using the pipeline (`%>%`) operator and piping our data into the `datatable()` function.

```{r create2}
station_data %>%
  datatable()
```

We immediately have an interactive table. The remainder of this tutorial will address how to tweak the table we have. Currently here are some of the things we can do:

- View our data in an aesthetic table
- Show 10, 25, 50 or 100 observations at a given time using the drop-down menu
- View the another page of observations with the page feature below the table
- Click on the arrows next to the column names to order our observations in ascending or descending order by that variable name
- Search with the search bar for the occurence of a string in any column of an observation
    - Try typing "Tarcoola"
    - Now try typing "25". What do you notice?
    
### Additional Options

We notice that the first 'column' is just a number representing which observation of the table we are looking at. This is generally of no purpose to us, so we can remove it using the `rownames` argument of `datatable()`. We set it to `FALSE`.

```{r improve1}
station_data %>%
  datatable(rownames = FALSE)
```

We also notice some of our column names are not very aesthetic. We can make them more so by:

- Using the [`gsub()`](http://www.endmemo.com/program/R/gsub.php) and `stringr::str_to_title()` functions to create a list of nicer names
- Using the `colnames` argument of `datatable()` to change these names

The following command replaces all dots (`.`s) with spaces, then capitalises the first letter of each word, and stores the new set of names in the `column_title` variable.

```{r improve2}
column_titles <- gsub("[.]", " ", colnames(station_data)) %>%
  stringr::str_to_title()

column_titles
```

We now use these as the new column names with `colnames`:

```{r improve3}
station_data %>%
  datatable(rownames = FALSE,
            colnames = column_titles)
```

We can already search for a given term in our data, but if we want to search for a term in a single column, we can add individual column filters using the `filter` argument.

- This argument takes a list of sub-arguments as its value
- We set the `position` sub-argument to "top" or "bottom" to add our filter

```{r improve4}
station_data %>%
  datatable(rownames = FALSE,
            colnames = column_titles,
            filter = list(position = "top"))
```

If we wish to display a number of observations which is not 10, 25, 50 or 100, we can use the `options` argument,

- This argument takes a list of sub-arguments as its value
- We set the `pageLength` sub-argument to the number of observations we wish to display

```{r improve5}
station_data %>%
  datatable(rownames = FALSE,
            colnames = column_titles,
            options = list(pageLength = 7))
```

Now say we wish to modify the search bar so that it doesn't say "Search:", but instead "Keyword look-up:".

- We still use the `options` argument of `datatable()`
- We use the sub-argument `language` of `options`, which itself takes a list as its argument
- We use the sub-sub-argument `sSearch` to rename the search bar

```{r improve6}
station_data %>%
  datatable(rownames = FALSE,
            colnames = column_titles,
            options = list(pageLength = 7,
                           language = list(sSearch = "Keyword look-up:")))
```

## Formatting Values in Tables
> [LinkedIn Learning 6.3](https://www.linkedin.com/learning/r-interactive-visualizations-with-htmlwidgets/format-columns-and-data)

```{r sample1, echo=FALSE}
set.seed(10)
sample_price <- tibble(Date = c("21-06-2019",
                                "22-06-2019",
                                "23-06-2019",
                                "24-06-2019",
                                "25-06-2019",
                                "26-06-2019"),
                       Budget = c(50,
                                  50,
                                  50,
                                  50,
                                  50,
                                  50),
                       Amount_Spent = rgamma(6, 25, 1/2))
```

Sometimes we will wish to display data in specific formats, such as dollars (`$`), a percentage (`%`) or otherwise. We have specific functions in DT for each of these.

```{r format1}
sample_price %>%
  datatable()
```

If we wish to format a column as a currency, we pipe our table into `formatCurrency()`:

- The first argument is the column to format (as a string!!!)
- We then use the `currency` argument to select the currency we want (as a character)

```{r format2}
sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "$")
```

If we were dealing in another currency, such as the Japanese Yen, we can specify an alternate currency:

```{r format3}
sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "¥")
```

If we wish to change the decimal places displayed, we use the additional argument `digits`:

```{r format4}
sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "$") %>%
  formatCurrency("Budget",
                 currency = "$",
                 digits = 0)
```

Let's say we now want to display the percentage of our budget we have spent.

```{r format5}
sample_price <- sample_price %>%
  mutate(Portion_Spent = Amount_Spent / Budget)
sample_price
```

We use the `formatPercentage()` function, specifying:

- The column to format as the first argument
- The number of decimal places we require as the `digits` argument

```{r format6}
sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "$") %>%
  formatCurrency("Budget",
                 currency = "$",
                 digits = 0) %>%
  formatPercentage("Portion_Spent",
                 digits = 2)
```

Lastly, if we wish to format dates, we use the `formatDate()` function:

- We specify the column to format as the first argument
- We specify the method by which to format in the `method` column
    - We generally assign "toDateString", but [a full list of methods can be found here](https://rstudio.github.io/DT/functions.html)
    
```{r format7}
sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "$") %>%
  formatCurrency("Budget",
                 currency = "$",
                 digits = 0) %>%
  formatPercentage("Portion_Spent",
                 digits = 2) %>%
  formatDate("Date",
             method = "toDateString")
```

Here the date formating failed. The reason for this is that the "Date" column contains strings of D-M-Y formatted dates. The `formatDate()` function requires entries of class 'date', not strings. We can fix this using the `lubridate` package.

- We mutate our Date column
- We use the `dmy()` function of `lubridate` because our dates are in D-M-Y format
- The entries are reclassified as dates

```{r format8}
class(sample_price$Date)

library("lubridate")
sample_price <- sample_price %>%
  mutate(Date = dmy(Date))

class(sample_price$Date)

sample_price %>%
  datatable() %>%
  formatCurrency("Amount_Spent",
                 currency = "$") %>%
  formatCurrency("Budget",
                 currency = "$",
                 digits = 0) %>%
  formatPercentage("Portion_Spent",
                 digits = 2) %>%
  formatDate("Date",
             method = "toDateString")
```

## Options for Responsive Tables
> [LinkedIn Learning 6.4](https://www.linkedin.com/learning/r-interactive-visualizations-with-htmlwidgets/design-responsive-mobile-ready-tables)

If we have tables with lots of columns, we have options to truncate these.

(Sample data provided by [Climate Change in Australia](https://www.climatechangeinaustralia.gov.au/en/climate-projections/explore-data/data-download/station-data-download/))

```{r get_data, echo=FALSE}
fire_data <- readxl::read_xlsx("tidy_Fire_data/fire_summary_data.xlsx")
```

```{r response1}
fire_data %>%
  datatable(rownames = FALSE,
            options = list(pageLength = 5))
```

The simplest method is using the argument `extensions` of `datatable()`.

- We can set this to "Responsive" to truncate columns
- These can then be viewed for individual data points

```{r response2}
fire_data %>%
  datatable(rownames = FALSE,
            extensions = "Responsive",
            options = list(pageLength = 5))
```

We now have a `+` button that displays columns which can't fit on our screen. There is, however, an even more responsive solution which allows the reader to select and deselect columns for viewing as they like. This requires some setup:

- We set the `extensions` argument to a vector containing "Responsive" and "Buttons" (ie `c("Responsive", "Buttons"))
- We add the sub-argument `buttons` to the `options` argument of `datatable()`
- This takes the value of `I("colvis")` since we want the column-visability button
    - Note: the reason we wrap this in the `I()` function is technical and relates to the formatting of DT
- We add the sub-argument `dom` to the `options` argument of `datatable()`
- This takes the value of "Bf" for our purposes
    - More on `dom` below
    
```{r response3}
fire_data %>%
  datatable(rownames = FALSE,
            extensions = c("Responsive", "Buttons"),
            options = list(pageLength = 5,
                           buttons = I("colvis"),
                           dom = "Bf"))
```

We now have a "Column visibility" button we can use to toggle columns as displayed and hidden.

## More on the `dom` Argument

This argument controls much about what we see in our tables. The argument is always a string, but, unlike many arguments we have seen in R, it is the individual letters of the string that are important.

In the above example, "Bf" means we desire `B`uttons and `f`ilter to be enabled, ***in that order***. There are actually many letters we can use to enhance or change our table (several of which are described [here](https://datatables.net/reference/option/dom)). We may be interested in:

| Letter | Function |
| :-: | :-: |
| `l` | Enables length changing control |
| `f` | Enables the global filter |
| `t` | The table itself (**\***) |
| `i` | Table information summary |
| `p` | Page number navigation |
| `B` | Buttons (assuming they have been properly set up) |

(**\***) Of course the table is always enabled, but this option is important in controlling order. For example, if we want the filter to be _below_ the table, we use `dom = "tf"`:

```{r dom1}
station_data %>%
  datatable(rownames = FALSE,
            options = list(pageLength = 5,
                           dom = "tf"))
```

As another example, consider the argument "Bpiltf", structuring the widget, from top to bottom, as buttons, page-navigation, summary info, length changing, the table, and the filter below:

```{r dom2}
station_data %>%
  datatable(rownames = FALSE,
            extensions = c("Responsive", "Buttons"),
            options = list(pageLength = 5,
                           buttons = I("colvis"),
                           dom = "Bpiltf"))
```

Lastly, as a fun example, consider that we are not limited to just one of each letter:

```{r dom3}
station_data %>%
  datatable(rownames = FALSE,
            options = list(pageLength = 5,
                           dom = "ppfptpf"))
```

Maybe this would not be such a good idea in practice!

## Allowing Users to Download Table Data
> [LinkedIn Learning 6.5](https://www.linkedin.com/learning/r-interactive-visualizations-with-htmlwidgets/allow-users-to-download-data)

We can add a button which allows users to download the data as it appears in our table.

- Users may download the content straight from (for example) your HTML webpage
- Users also may filter or remove columns and download the *modified* table

To do this is very similar to how we added the "colvis" buttons:

- Include the `extensions` argument of `datatable()`
- Set this to `c("Responsive", "Buttons")`
- Include the `options` argument of `datatable()`
- Set this to a list, containing the sub-argument `buttons`
- Set the value of this sub-argument to a vector of "excel" and "csv"

```{r download1}
station_data %>%
  datatable(rownames = FALSE,
            extensions = c("Responsive", "Buttons"),
            options = list(pageLength = 5,
                           buttons = c("colvis", "excel", "csv"),
                           dom = "Bf"))
```